Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
fixed units

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid \ttt\mid\fff\mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph with legal binders. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    \multicolumn{4}{c}{
      \quad\graphof{\ttt}=\single\ttt
      \qquad
      \graphof{\fff}=\single\fff
      \qquad
    \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

Two atoms are \bfit{pre-dual} if their predicate symbols are dual
(e.g. $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} $\tuple{\gC,\linkingof\gC}$ is a coloured fograph $\gC$ such
that every colour (i.e., equivalence class of~$\linkingof\gC$), called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link.

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{A}{
        \vlde{\set{\acD,\cfaD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}
\noindent
This theorem is stronger than the existing decompositions for
first-order logic, which either separate only atomic contraction and
atomic weakening~\cite{brunnler:phd} or only
contraction~\cite{ralph:phd} or only the quantifiers in form of a
Herbrand theorem~\cite{brunnler:06:locality,ralph:phd}.

@@

\begin{example}
  Below is an example of a decomposed derivation in $\FOKS$ of the
  formula $(\exists x. \cneg{p}(x)) \vlor (\forall y.(p(y) \vlan
  p(f(y))))$:
  \begin{equation*}
    \vlderivation{
      \vlin{\acD}{}{(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))}{
        \vlin{\mexD}{}{(\exists x. (\cneg{p}x \vlor \cneg{p}x) \vlor (\forall
          y.(py \vlan pfy))}{
          \vlin{\fequ}{}{((\exists x. \cneg{p}x) \vlor (\exists x.\cneg{p}x))
            \vlor (\forall y.(py \vlan pfy)}{
            \vlin{\exists}{}{\forall y.(((\exists x. \cneg{p}x) \vlor (\exists
              x.\cneg{p}x)) \vlor (py \vlan pfy))}{
              \vlin{\exists}{}{\forall y.((\cneg{p}y \vlor (\exists x.\cneg{p}x))
                \vlor (py \vlan pfy))}{
                \vlin{\fequ}{}{\forall y.((\cneg{p}y \vlor \cneg{p}fy) \vlor
                  (py \vlan pfy))}{
              \vlin{\sw}{}{\forall y.(\cneg{p}y \vlor ((py \vlan pfy)
                \vlor \cneg{p}fy))}{
                \vlin{\fequ}{}{\forall y.(\cneg{p}y \vlor (py \vlan (pfy
                  \vlor \cneg{p}fy)))}{
                  \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan (pfy
                    \vlor \cneg{p}fy))}{
                    \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan
                      \ttt)}{
                      \vlin{\ttt}{}{\forall y.(\ttt \vlan \ttt)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
  \end{equation*}
\end{example}
%
There is a weaker version of Theorem~\ref{thm:decomposition} that will
also be useful:
%
\begin{thm}\label{thm:decompositionA}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there is a formula $A_1$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:CP-DI}
  Let  $\phi\colon\gC\to\gA$ be a combinatorial proof and let $A$ be a formula with $\gA=\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'\fequ C\substof\phi$ where $C$ is a formula with $\graphof
  C=\gC$ and $\substof\phi$ is the variable renaming substitution
  induced by~$\phi$.  Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\begin{proof}
  This is a straightforward induction on the proof of $\sqn\Gamma$ in
  $\FOMLL$, making a case analysis on the bottommost rule instance. We
  show here only the case of $\vlinf{\forall}{}{\sqn{\Delta,\forall
      x.A}}{\sqn{\Delta,A}}$ (all other cases are simpler and have
  been shown before, e.g.~\cite{brunnler:phd}): By induction
  hypothesis, there is a proof of $\form\Delta\vlor A$ in $\FOMLS$. We
  can prefix every line in that proof by $\forall x$ and then compose
  the following derivation:
  \begin{center}\vspace{-2ex}\begin{math}
    \vlderivation{
      \vlin{\fequ}{}{\form\Delta\vlor\forall x.A}{
        \vlde{\FOMLS}{}{\forall x.\form\Delta\vlor A}{
          \vlin{\forall}{}{\forall x.\ttt}{
            \vlhy{\ttt}}}}}
  \end{math}\end{center}
  where we can apply the $\fequ$-rule because $x$ is not free in $\Delta$.
\end{proof}

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$ other than $\aiD$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\begin{proof}
  This is a straightforward exercise that we leave to the reader. (Note that the
$\ax$-rule can be applied to $\sqn{\fff,\ttt}$ in the cases of $\rr = \faD$.)
\end{proof}

@@

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof $\Proof$ of $\sqn{\Scons A}$. If $\rr$ is
  $\upsmash{\vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}}$, we replace
  in $\Pi$ all corresponding occurrences of $\ttt$ with
  $a \vlor \cneg a$ and the rule instance
  $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation $ \vlderivation{
    \vlin{\vlor}{}{\sqn{a \vlor \cneg a}}{ \vlin{\axr}{}{\sqn{a, \cneg
          a}}{ \vlhy{}}}} $. This gives us a proof of
  $\sqn{\Scons{a \vlor \cneg a}}$. In all other cases, 
    %\item $\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$
     % In this case, we replace simultaneously in $\Pi$ the corresponding
     % occurrences of $\ttt$ with $\forall x.\ttt$ and the rule instance
     % $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
     % $
     % \vlderivation{
     %   \vlin{\forall}{}{\sqn{\forall x.\ttt}}{
     %     \vlin{\axr}{}{\sqn{\ttt}}{
     %       \vlhy{}}}}
     % $, and we get a proof $\Pi'$ of $\sqn{\Scons{\forall x.\ttt}}$.
      by Lemmas~\ref{lem:shallow} and~\ref{lem:context}, we have a $\FOMLL$ proof of
      $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
      \begin{equation*}
        \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
      \end{equation*}
      and then apply Theorem~\ref{thm:cutelim}.
  \end{proof}

@@

\begin{proof}
  We show that there is always a derivation
  $$
  \vlderivation{
    \vlde{\sysS}{}{A}{
      \vlhy{A \vlor A}}}
  $$, where $\sysS = \set{\acD,\me,\mfaD,\mexD,\fequ}$, by induction on $A$:
  \begin{itemize}
    \item If $A = a$, then we have
      $
      \vlderivation{
        \vlin{\acD}{}{a}{
          \vlhy{a \vlor a}}}
      $.
    \item If $A = B \vlan C$, then we have
      $
      \vlderivation{
	\vlde{\sysS}{}{B \vlan C}{
	  \vlde{\sysS}{}{B \vlan (C \vlor C)}{
	    \vlin{\me}{}{(B \vlor B) \vlan (C \vlor C)}{
	      \vlhy{(B \vlan C) \vlor (B \vlan C)}}}}}
      $.\\[.5ex]
    \item If $A = B \vlor C$, then we have
      $
      \vlderivation{
	\vlde{\sysS}{}{B \vlor C}{
	  \vlde{\sysS}{}{B \vlor (C \vlor C)}{
	    \vlin{\fequ}{}{(B \vlor B) \vlor (C \vlor C)}{
	      \vlhy{(B \vlor C) \vlor (B \vlor C)}}}}}
      $.\\[.5ex]
    \item If $A = \exists x.A'$, then we have
      $
      \vlderivation{
	\vlde{\sysS}{}{\exists x.A'}{
	  \vlin{\mexD}{}{\exists x.(A' \vlor A')}{
	    \vlhy{(\exists x.A') \vlor (\exists x.A')}}}}
      $.\\[.5ex]
    \item If $A = \forall x.A'$, then we have
      $
      \vlderivation{
	\vlde{\sysS}{}{\forall x.A'}{
	  \vlin{\mexD}{}{\forall x.(A' \vlor A')}{
	    \vlhy{(\forall x.A') \vlor (\forall x.A')}}}}
      $.
  \end{itemize}
  \todo{}
  \juihsuan{done. Maybe just keep one case.}\lutz{yes, but we do that at the end. don't think about space right now.}
\end{proof}

@@

\begin{proof}
  \todo{}

>>>>

Fix pairwise disjoint countable sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability of often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph with legal binders. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

Two atoms are \bfit{pre-dual} if they are not $\ttt$ or $\fff$, and their predicate symbols are dual
(e.g. $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} $\tuple{\gC,\linkingof\gC}$ is a coloured fograph $\gC$ such
that every colour (i.e., equivalence class of~$\linkingof\gC$), called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link. Hence, in a linked fograph no vertex is labeled~$\fff$.

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are $\fff$-free formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\fequ}}{}{A}{
        \vlde{\set{\acD,\cfaD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
    $}
  \end{equation*}
\end{thm}
\noindent
This theorem is stronger than the existing decompositions for
first-order logic, which either separate only atomic contraction and
atomic weakening~\cite{brunnler:phd} or only
contraction~\cite{ralph:phd} or only the quantifiers in form of a
Herbrand theorem~\cite{brunnler:06:locality,ralph:phd}.

@@

%\begin{example}
  Figure~\ref{fig:example-decompose} shows an example of a decomposed derivation in $\FOKS$ of the
  formula $(\exists x. \cneg{p}(x)) \vlor (\forall y.(p(y) \vlan
  p(f(y))))$.
%\end{example}

\begin{figure}
  \begin{equation*}
    \scalebox{.9}{$
    \vlderivation{
      \vlin{\acD}{}{(\exists x. \cneg{p}x) \vlor (\forall y.(py \vlan pfy))}{
        \vlin{\mexD}{}{(\exists x. (\cneg{p}x \vlor \cneg{p}x) \vlor (\forall
          y.(py \vlan pfy))}{
          \vlin{\fequ}{}{((\exists x. \cneg{p}x) \vlor (\exists x.\cneg{p}x))
            \vlor (\forall y.(py \vlan pfy)}{
            \vlin{\exists}{}{\forall y.(((\exists x. \cneg{p}x) \vlor (\exists
              x.\cneg{p}x)) \vlor (py \vlan pfy))}{
              \vlin{\exists}{}{\forall y.((\cneg{p}y \vlor (\exists x.\cneg{p}x))
                \vlor (py \vlan pfy))}{
                \vlin{\fequ}{}{\forall y.((\cneg{p}y \vlor \cneg{p}fy) \vlor
                  (py \vlan pfy))}{
              \vlin{\sw}{}{\forall y.(\cneg{p}y \vlor ((py \vlan pfy)
                \vlor \cneg{p}fy))}{
                \vlin{\fequ}{}{\forall y.(\cneg{p}y \vlor (py \vlan (pfy
                  \vlor \cneg{p}fy)))}{
                  \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan (pfy
                    \vlor \cneg{p}fy))}{
                    \vlin{\aiD}{}{\forall y.((\cneg{p}y \vlor py) \vlan
                      \ttt)}{
                      \vlin{\ttt}{}{\forall y.(\ttt \vlan \ttt)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
    $}
  \end{equation*}    
    \caption{Example derivation in decomposed form of Theorem~\ref{thm:decomposition}}
    \label{fig:example-decompose}
  \end{figure}
%
There is a weaker version of Theorem~\ref{thm:decomposition} that will
also be useful:
%
\begin{thm}\label{thm:decompositionA}
  For every derivation $\upsmash{\vlder{\FOKS}{\Deri}{A}{\ttt}}$ there is a formula~$A_1$ not containing any occurence of $\fff$, such that there is a derivation:
  \begin{equation*}
    %\scalebox{.9}{$
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}
   % $}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:CP-DI}
  Let  $\phi\colon\gC\to\gA$ be a combinatorial proof and let $A$ be a formula with $\gA=\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\wfaD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'\fequ C\rsubstof\phi$ where $C$ is a formula with $\graphof
  C=\gC$ and $\rsubstof\phi$ is the variable renaming substitution
  induced by~$\phi$.  Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, such that $\fff$ does not occur in~$A_1$, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\begin{proof}
  This is a straightforward induction on the proof of $\sqn\Gamma$ in
  $\FOMLL$, making a case analysis on the bottommost rule instance. We
  show here only the case of
%%  $\downsmash{\vlinf{\fff}{}{\sqn{\Delta,\fff}}{\sqn\Delta}}$ and
  $\vlinf{\forall}{}{\sqn{\Delta,\forall x.A}}{\sqn{\Delta,A}}$ (all
  other cases are simpler or have been shown before,
  e.g.~\cite{brunnler:phd}): By induction hypothesis, there is a proof
  of $\form\Delta\vlor A$ in $\FOMLS$. We can prefix every line in
  that proof by $\forall x$ and then compose the following derivation:
  \begin{center}\vspace{-2ex}\begin{math}
    \vlderivation{
      \vlin{\fequ}{}{\form\Delta\vlor\forall x.A}{
        \vlde{\FOMLS}{}{\forall x.\form\Delta\vlor A}{
          \vlin{\forall}{}{\forall x.\ttt}{
            \vlhy{\ttt}}}}}
  \end{math}\end{center}
  where we can apply the $\fequ$-rule because $x$ is not free in $\Delta$.
\end{proof}

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\begin{proof}
  This is a straightforward exercise.
%%  that we leave to the reader. (Note that the
%%$\ax$-rule can be applied to $\sqn{\fff,\ttt}$ in the cases of $\rr = \faD$.)
\end{proof}

@@

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof $\Proof$ of $\sqn{\Scons A}$.
  %% If $\rr$ is
  %% $\upsmash{\vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}}$, we replace
  %% in $\Pi$ all corresponding occurrences of $\ttt$ with
  %% $a \vlor \cneg a$ and the rule instance
  %% $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation $ \vlderivation{
  %%   \vlin{\vlor}{}{\sqn{a \vlor \cneg a}}{ \vlin{\axr}{}{\sqn{a, \cneg
  %%         a}}{ \vlhy{}}}} $. This gives us a proof of
  %% $\sqn{\Scons{a \vlor \cneg a}}$. In all other cases, 
  %%   %\item $\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$
  %%    % In this case, we replace simultaneously in $\Pi$ the corresponding
  %%    % occurrences of $\ttt$ with $\forall x.\ttt$ and the rule instance
  %%    % $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
  %%    % $
  %%    % \vlderivation{
  %%    %   \vlin{\forall}{}{\sqn{\forall x.\ttt}}{
  %%    %     \vlin{\axr}{}{\sqn{\ttt}}{
  %%    %       \vlhy{}}}}
  %%    % $, and we get a proof $\Pi'$ of $\sqn{\Scons{\forall x.\ttt}}$.
      By Lemmas~\ref{lem:shallow} and~\ref{lem:context}, we have a $\FOMLL$ proof of
      $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
      \begin{equation*}
        \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
      \end{equation*}
      and then apply Theorem~\ref{thm:cutelim}.
  \end{proof}

@@

\begin{proof}
  This can be shown by a straightforward induction on $A$ (for details, see e.g.~\cite{brunnler:phd}).
  %% We show that there is always a derivation
  %% $$
  %% \vlderivation{
  %%   \vlde{\sysS}{}{A}{
  %%     \vlhy{A \vlor A}}}
  %% $$, where $\sysS = \set{\acD,\me,\mfaD,\mexD,\fequ}$, by induction on $A$:
  %% \begin{itemize}
  %%   \item If $A = a$, then we have
  %%     $
  %%     \vlderivation{
  %%       \vlin{\acD}{}{a}{
  %%         \vlhy{a \vlor a}}}
  %%     $.
  %%   \item If $A = B \vlan C$, then we have
  %%     $
  %%     \vlderivation{
  %%       \vlde{\sysS}{}{B \vlan C}{
  %%         \vlde{\sysS}{}{B \vlan (C \vlor C)}{
  %%           \vlin{\me}{}{(B \vlor B) \vlan (C \vlor C)}{
  %%             \vlhy{(B \vlan C) \vlor (B \vlan C)}}}}}
  %%     $.\\[.5ex]
  %%   \item If $A = B \vlor C$, then we have
  %%     $
  %%     \vlderivation{
  %%       \vlde{\sysS}{}{B \vlor C}{
  %%         \vlde{\sysS}{}{B \vlor (C \vlor C)}{
  %%           \vlin{\fequ}{}{(B \vlor B) \vlor (C \vlor C)}{
  %%             \vlhy{(B \vlor C) \vlor (B \vlor C)}}}}}
  %%     $.\\[.5ex]
  %%   \item If $A = \exists x.A'$, then we have
  %%     $
  %%     \vlderivation{
  %%       \vlde{\sysS}{}{\exists x.A'}{
  %%         \vlin{\mexD}{}{\exists x.(A' \vlor A')}{
  %%           \vlhy{(\exists x.A') \vlor (\exists x.A')}}}}
  %%     $.\\[.5ex]
  %%   \item If $A = \forall x.A'$, then we have
  %%     $
  %%     \vlderivation{
  %%       \vlde{\sysS}{}{\forall x.A'}{
  %%         \vlin{\mexD}{}{\forall x.(A' \vlor A')}{
  %%           \vlhy{(\forall x.A') \vlor (\forall x.A')}}}}
  %%     $.
  %% \end{itemize}
  %% \todo{}
  %% \juihsuan{done. Maybe just keep one case.}\lutz{yes, but we do that at the end. don't think about space right now.}
\end{proof}

@@

\begin{proof}
  We only show the cases for $\wfaD$ and $\cfaD$:
  $$
  \todo{}
  \vlderivation{
    \vlin{\cD}{}{\forall x.A}{
      \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
        \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
          \vlhy{\forall x.\forall x.A}}}}}
  $$
  The other cases have been shown before (e.g.~\cite{brunnler:phd}).
%%   We have the following derivations:
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.A}{
%%       \vlin{\equiv}{(x \notin fv(\forall x.A))}{(\forall x.A) \vlor (\forall x.A)}{
%%         \vlin{\wrD}{}{\forall x.((\forall x.A) \vlor A)}{
%%           \vlhy{\forall x.\forall x.A}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{(A \vlor B) \vlan (C \vlor D)}{
%%       \vlin{\fequ}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((A \vlor B) \vlan (C \vlor D))}{
%%       \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan (D \vlor C))}{
%% 	\vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor ((B \vlor A) \vlan D))}{
%% 	  \vlin{\wrD}{}{((A \vlor B) \vlan (C \vlor D)) \vlor (B \vlan D))}{
%% 	    \vlin{\wrD}{}{((A \vlor B) \vlan C) \vlor (B \vlan D)}{
%% 	      \vlhy{(A \vlan C) \vlor (B \vlan D)}}}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\exists x.(A \vlor B)}{
%%       \vlin{\fequ}{}{\exists x.(A \vlor B)) \vlor (\exists x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\exists x.(A \vlor B)) \vlor (\exists x.B)}{
%% 	  \vlhy{(\exists x.A) \vlor (\exists x.B)}}}}}}
%%   $$.
%%   $$
%%   \vlderivation{
%%     \vlin{\cD}{}{\forall x.(A \vlor B)}{
%%       \vlin{\fequ}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(A \vlor B))}{
%%       \vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.(B \vlor A))}{
%% 	\vlin{\wrD}{}{(\forall x.(A \vlor B)) \vlor (\forall x.B)}{
%% 	  \vlhy{(\forall x.A) \vlor (\forall x.B)}}}}}}
%%   $$.
%%   \juihsuan{done. If needed, we can introduce the notion of open deduction to
%% reduce the size of derivations...}\lutz{I was thinking about that, but (i) it is probably not worth the effort, as we wont have many derivation, and (ii) it is hard to define rectified derivations this way.}
\end{proof}
